How to reduce Item Nonresponse in Face-to-Face Surveys? A Review and Evidence from the European Social Survey

### 2024-01-31 inr_ESS9_text.tex
revised text after second round of reviews

%opening
\title{How to reduce Item Nonresponse in Face-to-Face Surveys? A Review and Evidence from the European Social Survey}
\author{Malte Grönemann \\ University of Mannheim, Mannheim, Germany}

@@

\begin{abstract}
I review the literature on item nonresponse in surveys and based on this review, I extend the satisficing model with respondents' privacy concerns to incorporate all relevant aspects of the response process for item nonresponse. I review proposed strategies to reduce item nonresponse and test selected strategies. Results suggest that boosting respondents' use of showcards, shorter interview durations and conducting the interview in the respondents' primary language might be promising ways to reduce item nonresponse. Other people present during the interview only has a small association with the number of refusals. Matching age and gender of respondents and interviewers appears not to be a worthwile strategy.
\end{abstract}

@@

Missing data pose a problem to the analysis of survey data. They decrease the effective sample size and can introduce bias to estimates if the causes for missingness are related to the item or respondent characteristics \citep{deleeuwPreventionTreatmentItem2003}. Although missing data are relatively rare in most single items, they can add up to a considerable loss of observations in multivariate analyses. Due to these harmful effects of missing data, one objective of survey researchers is to keep their prevalence as low as possible. Therefore, it is important to understand the processes that can lead to missing data.

I focus on the respondent as a source of missing data by answering with ``Don't know'' (DK) or refusing to answer. (\textit{item nonresponse}). Item nonresponse can also be seen as an indicator of overall data quality since it can be the result of \textit{satisficing} \citep{krosnickResponseStrategiesCoping1991}. Satisficing means that the respondent is giving a satisfactory answer instead of the best one possible.

I review the existing literature on item nonresponse and extend the satisficing model based on this review to include privacy concerns resulting in an encompassing model of item nonresponse: the probability of item nonresponse depends on the task difficulty of the item(s) divided by the product of ability and motivation of the respondent or the respondents' privacy concerns. This means that higher difficulty results in more item nonresponse, higher ability and motivation in less item nonresponse. Difficulty, ability and motivation are separate from privacy concerns, e.g. due to item sensitivity or general mistrust. Privacy concerns are relevant when deciding whether to disclose an already formed answer and does not influence the cognitive burden of retrieving the answer.

In the empirical part of the paper, I explore a few practical ideas to reduce item nonresponse. Promoting the use of showcards, shorter interviews and tranlating questionnaires might be promising strategies. All of those reduce the cognitive burden of respondents. The effects are not causal though because I am not able to sufficiently control for respondent ability. Matching respondents' and interviewers' gender and age seem not to be efficient tools to reduce item nonresponse. They could have influenced item nonresponse if respondents are more willing to share private information with interviewers that are similar to them.

This introduction follows a short theoretical discussion presenting a model for the probability of item nonresponse based on satisficing. Structured by this model, I review the literature on strategies how to reduce item nonresponse and test a selection using the \textit{European Social Survey}. The rest of the paper is devoted to this test of strategies, describing data and methods and presenting results. Finally, I summarise and discuss my review and results.

@@

The \textit{survey response process} \citep{tourangeauPsychologySurveyResponse2000} involves multiple steps on behalf of the respondent. They need to comprehend the question, retrieve information from memory, eventually they need to judge these information, map them onto the response options and perhaps edit the response due to sensitivity or social desirability. Respondents can take these steps in order or jump back and forth, e.g. if they need to form an opinion on the spot. But in all of these steps, item nonresponse can be introduced \citep{deleeuwPreventionTreatmentItem2003}.

The two types of item nonresponse might be related to different steps of the survey response process. Refusals are more likely to stem from the editing step when respondents do not want to answer although they could. They might feel an item is too sensitive or do not trust the interviewer with this information. DK is more likely to come about when the respondent cannot answer in the first place. Either the respondent cannot answer because they do not know about the content of the question or are unable to remember an event \citep{beattyAnswerNotAnwer2002, turnerCanResponseLatencies2015}. Or they do not perceive it worth the effort to come up with an answer and satisfice \citep{krosnickResponseStrategiesCoping1991}. But \citet{shoemakerItemNonresponseDistinguishing2002} have shown that higher mental effort is related to more refusals as well. And conversely, it is plausible that DK is used as a more polite way to refuse. I, therefore, do not distinguish between them in the following sections even though they often show different strengths of predictors \citep{silberEffectsQuestionRespondent2021}.

Ability has been the explanation for the differences often found in item nonresponse by education, age, physical and mental health \citep{colsherDataQualityAge1989, pickeryImpactRespondentInterviewer1998, deleeuwPreventionTreatmentItem2003, messerDeterminantsItemNonresponse2012, silberEffectsQuestionRespondent2021}. Women and ethnic minority members are often found to have more item nonresponse \citep{kupekDeterminantsItemNonresponse1998, pickeryImpactRespondentInterviewer1998}. For the latter, on average lower literacy and lower command of the majority language are hypothesised to cause more item nonresponse \citep{kupekDeterminantsItemNonresponse1998}. \citet{meitingerPowerCultureItem2020} conclude that item nonresponse reflects broader social inequalities in abilities and access to information. The ability hypothesis is also directly supported by correlations between item nonresponse and measures of intelligence \citep{hedengrenDogThatDidn2012}.

More difficult or unclear questions typically have higher levels of item nonresponse \citep{holbrookImpactQuestionRespondent2006, messerDeterminantsItemNonresponse2012, holbrookInterviewerErrorsHelp2016, olsonEffectsRespondentQuestion2019}. Since demographic questions can often be recalled very simply by respondents, they often show less item nonresponse than attitudinal and behavioural questions where the answers often do not exist before before getting asked the question \citep{olsonEffectsRespondentQuestion2019, silberEffectsQuestionRespondent2021}.

Even when able to perform the task at hand, respondents not necessarily want to.
They need to perceive the effort as worth it. Differences in \textit{motivation} are an explanation
for less item nonresponse of respondents that are more interested in the topic of the survey \citep{kochItemNonresponseEuropean2009, silberEffectsQuestionRespondent2021}. Item nonresponse is related to measures of conscientiousness as well \citep{hedengrenDogThatDidn2012}.

Motivation can also affect the editing process. For example, respondents (and interviewers) might deliberately not answer screening and filtering questions to reduce survey burden \citep{tourangeauMotivatedMisreportingShaping2015}.
When editing an answer, concerns of privacy\footnote{I use this label to subsume general concerns about privacy towards the survey (e.g. about data processing, anonymity) and not wanting to answer single sensitive questions.} come into play. Am I going to be judged by the interviewer when answering honestly? Are my data really secure and confidential? This is a particular problem for questions perceived as more intrusive \citep{tourangeauSensitiveQuestionsSurveys2007}. Most prominently, this affects questions on income \citep{yanTrendsIncomeNonresponse2010} and sexual behaviour \citep{kupekDeterminantsItemNonresponse1998}, which often show particularly high levels of item nonresponse. High item nonresponse is often used as an indicator for question sensitivity \citep{tourangeauSensitiveQuestionsSurveys2007}. Reluctancy and overall skepticism towards surveys and science, privacy concerns and generalised mistrust are traits that result in more nonresponse \citep{silberEffectsQuestionRespondent2021}.

@@

I have thus identified four constructs related to the probability of item nonresponse under which most research can be subsumed: cognitive ability, task difficulty, motivation and privacy concerns. However, as \citet{krosnickResponseStrategiesCoping1991} already hypothesised, these concepts are interrelated in their effect on item nonresponse. Very easy questions can be answered by less able respondents and very hard questions might even cause the most able to struggle. The resulting fraction of difficulty by ability represents the relative mental effort to answer a question. To what point the relative mental effort is worth it depends on the respondents' motivation. A highly motivated respondent even answeres the difficult questions. The probability of nonresponse decreases with motivation. \citet[225]{krosnickResponseStrategiesCoping1991} has summarised these relations into a formula of the probability of satisficing. Satisficing refers to respondents lowering their standards in answering a question and only giving a satisfactory instead of an optimal answer to cut down on cognitive effort. Item nonresponse is such a satisficing strategy. There are situations where DKs are genuine though \citep{turnerCanResponseLatencies2015}. This is nonetheless a product of the respondents' ability and thereby included in the formula.

Additionally, higher privacy concerns lead to more item nonresponse. Since this relates to another step in the survey response process, namely editing rather than comprehension, retrieval or judging, I postulate it to be independent from the other concepts. I therefore suggest extending the formula by Krosnick as follows:

$$ P_{INR} \ = \ max \left( \frac{difficulty}{ability \ * \ motivation} \ , \ privacy \right) $$

\textit{The probability of a nonsubstantive answer by the respondent is given by the task difficulty
divided by the product of ability and motivation or the respondents' privacy concerns, whichever
is higher.}

This formula specifically addresses item nonresponse for a single item but its meaning is adaptable to every level of a survey.

@@

Based on these four concepts, one can theorise how to decrease item nonresponse. Some of the following possibilities may seem obvious and are established standards not only because of their potential relationship to item nonresponse but to ensure quality of substantial answers as well. Others might reduce item nonresponse but they could have negative consequences for other parts of \textit{total survey error}, the combined effect of all error sources in a survey \citep{grovesTotalSurveyError2010}. They require a trade-off before implementation.

I have structured this review of approaches to reduce item nonresponse by the respective concepts they target.

@@

Task difficulty is predominantly influenced by design choices regarding single questions and their type of expected response. For the design of questions, much typical advice from question design aims to reduce task difficulty: asking short questions with simple structures, avoiding ambiguity and fuzzy concepts and not asking respondents to do multiple things at once. For a more comprehensive review on advice on question and questionnaire design, see for example \citet{smythDesigningQuestionsQuestionnaires2016}. But the type of question is also associated with task difficulty: open-ended questions, being allowed to select multiple options, more response categories and being asked to order categories are associated with higher item nonresponse \citep{schumanOpenClosedQuestion1979, holbrookImpactQuestionRespondent2006, holbrookInterviewerErrorsHelp2016, olsonEffectsRespondentQuestion2019, silberEffectsQuestionRespondent2021}. Additional measures to decrease difficulty can be visual helpers like images or showcards. With showcards, respondents do not need to remember all response categories while thinking about the question. But there is not much empirical work on the effect of showcards on item nonresponse. \citet{holbrookInterviewerErrorsHelp2016} found that questions using showcards had more item nonresponse although this is likely because they are only used with more difficult questions. In the context of the ESS, showcards do not seem to affect the distribution of substantive responses in experiments \citep{jackleAssessingEffectData2010} but item nonresponse was not investigated.

The question level is very well understood and constitutes the largest part of the variance in item nonresponse. This highlights the importance of the single question for overall data quality \citep{olsonEffectsRespondentQuestion2019}.

But not only single questions but the whole questionnaire needs to be designed as simply
as possible. Questionnaires with changes in response categories, routing and filtering are associated with higher nonresponse \citep{messerDeterminantsItemNonresponse2012}. Grouping questions by topic could reduce mental effort and therefore item nonresponse but it also increases the likelihood of non-differentiation between the items \citep{krosnickResponseStrategiesCoping1991}. Offering DK and refusal explicitly increases their use because respondents are more aware of the “easy way out” \citep{schumanAssessmentNoOpinion1979, beattyAnswerNotAnwer2002}.

Task difficulty can be reduced for language minorities by providing the questionnaire in more languages. However, this is expensive and can reduce comparability of cases. For a review on comparability in cross-cultural surveys, see e.g. \citet{behrTranslationMeasurementInstruments2016}.

@@

To make sure that the respondent is able to answer the questions, interviews should be conducted in situations where understanding and concentration is as good as possible. Other people present during the interview, for example, could be a source of distraction. Empirical investigations have not been conclusive though \citep{kupekDeterminantsItemNonresponse1998, tuSocialDistanceRespondent2007, silberEffectsQuestionRespondent2021}. With longer interviews, respondents might get tired \citep{holbrookInterviewerErrorsHelp2016, olsonEffectsRespondentQuestion2019}.

@@

Motivation could decrease over the course of the interview as well. Different page layouts and progress bars might help with that in web surveys but evidence is mixed \citep{peytchevWebSurveyDesign2006, yanShouldStayShould2011, sarrafSurveyPageLength2014}. The
literature on cooperation enhancement, e.g. using incentives, has focussed on unit nonresponse. Some of these ideas might be applicable to increase respondent motivation within the interview as well. After all, unit and item nonresponse are linked: respondents that initially refused to participate have higher levels of item nonresponse \citep{yanRelationUnitNonresponse2010, frickerExaminingRelationshipNonresponse2010}.

@@

To tackle reluctancy to answer due to privacy concerns, survey researchers need to make sure that the respondents' data are secure and anonymous and communicate that. Explaining why the data are collected, how they are processed and what is done to protect their privacy is also ethically advisable and often legally required.

In face-to-face surveys, establishing trust depends on the respondent-interviewer dynamics. Based on the hypothesis that respondents might perceive interviewers similar to themselves as more trustworthy, scholars have assessed the impact of matching their characteristics. \citet{vercruyssenEffectSociodemographicMis2017} find less item nonresponse when matching age
of interviewers and respondents and matching gender reduces item nonresponse for males and increases it for females. Matching education has no effect \citep{vercruyssenEffectSociodemographicMis2017, silberEffectsQuestionRespondent2021}. \citet{tuSocialDistanceRespondent2007} on the other hand find age and education matching are potent predictors of item nonresponse.

@@

In this review, I have stressed the role of the respondent and how design choices can increase the probability of a substantial answer. But in face-to-face surveys, the interviewer can be a source of item nonresponse as well. This is left for further research. I proceed with testing a few proposed strategies for item nonresponse reduction based on this review.

Over time, respondents might lose the ability to concentrate and motivation. I therefore expect more item nonresponse, the longer the interview takes. If this is true, making questionnaires only as long as necessary is a strategy to reduce item nonresponse and maintain high data quality. However, the duration of the interview might also be dependent on the respondents' ability. Overall ability of the respondent needs to be controlled for in the model.

Interferences and other people being present during the interview are expected to increase item nonresponse because they can be a source of distraction and respondents might not want to answer some questions in front of people they know. If true, interviews should be conducted without others present if possible.

Showcards are intended to make the process of answering less burdensome because respondents do not need to remember the response options. If this has an effect on item nonresponse, showcards should be provided and their use promoted to ensure a high data quality. In this operationalisation, the interviewer assesses afterwards to what extent the respondent used the showcards. Respondent ability might influence how well they understand the procedures, ability needs to be controlled for again. The interviewer assessment is likely influenced by the overall impression the interviewer has from the interview and is therefore endogenous. The overall impression of the interviewer needs to be controlled for as well.

Language difficulties are likely to inhibit the response process. For respondents that primarily speak a different language at home than the interview is conducted in, the task might be more difficult. To ensure data quality from these respondents, translating the questionnaire is an option (although costly and comes with problems of comparability). Respondents for whom this is true are likely migrant groups that have lower levels of education than the native population. Therefore, ability needs to be controlled for.

Previous research suggests that respondents might be more willing to answer to socially similar interviewers. I test this proposition for gender and age. If true and the sampling frame contains some demographic information, survey agencies could assign interviewers based on these demographics. Whether the gender of respondent and interviewer matches can be seen as random while the probability of the respondent being older or younger than the interviewer is dependent on the respondents' age, which is correlated with ability. Therefore, respondent age needs to be controlled for \footnote{Interviewer age is accounted for in the fixed effects.}.

Table \ref{table_strat} summarises the strategies to be tested. Necessary controls are respondents' ability and specifically age and the interviewers' overall assessment of the interview. Based on the theory outlined above, I do not see that inclusion of any of these control variables or the other strategies are likely to distort the effect of another variable of interest. Therefore, I am going to test all effects in a single model.

\begin{table}[h]
\caption{Reduction Strategies to be Tested}
\centering
\begin{tabular}{| l l l |}
\hline
Strategy & Mechanism & Expectation \\
\hline
Duration of the Interview & Ability, Motivation & positive \\
Interference of the Interview & Ability, Privacy & positive \\
Use of Showcards & Difficulty & negative \\
Interview not primary Language & Difficulty & positive \\
Gender Matching & Privacy & negative \\
Interviewer more than 10 years older & Privacy & positive \\
Respondent more than 10 years older & Privacy & positive \\
\hline
\end{tabular}
\label{table_strat}
\end{table}

@@

Following this review and synthesis, I want to test the effectiveness of some potential strategies to reduce item nonresponse. I use the \textit{European Social Survey (ESS)} Round 9 collected in 2018 \citep{essericEuropeanSocialSurvey2019}. The ESS is a biannual face-to-face survey on attitudes and beliefs towards social and political topics in Europe. The data release 3.1 includes data from 49,519 respondents from 29 countries.

@@

The three dependent variables are the sum of DK, the sum of refusals and the total sum of item nonresponse for every respondent. Item nonresponse is counted only in variables that are presented to all respondents, i.e. that are not affected by filtering questions. DK and refusal are not presented as response options in the ESS but are explicitly recorded as such. Interviewers are instructed to accept them without probe.

\subsection{Independent Variables}

The independent variables of interest are discussed in the previous section.

>>>>

%opening
\title{How to reduce Item Nonresponse in Face-to-Face Surveys? A Review and Evidence from the European Social Survey}
%\author{Malte Grönemann \\ University of Mannheim, Mannheim, Germany}

@@

\begin{abstract}
I review the literature on item nonresponse in surveys. Based on this review, I extend the satisficing model with respondents' privacy concerns to incorporate all relevant aspects of the response process for item nonresponse. I review proposed strategies to reduce item nonresponse and test selected strategies. Results suggest that boosting respondents' use of showcards and interviewing in the respondents' primary language might be promising ways to reduce item nonresponse. Other people present during the interview have only a small association with the number of refusals. Matching the age and gender of respondents and interviewers appears not to be a worthwhile strategy.
\end{abstract}

@@

Missing data pose a problem to the analysis of survey data. They decrease the effective sample size and can introduce bias to estimates if the causes for missingness are related to the item or respondent characteristics \citep{deleeuwPreventionTreatmentItem2003}. Although missing data are rare in most single items, they can add up to a considerable loss of observations in multivariate analyses. Item nonresponse can also be seen as an indicator of overall data quality since it can result from \textit{satisficing} \citep{krosnickResponseStrategiesCoping1991}. Satisficing means that the respondent is giving a satisfactory answer instead of the best one. Due to these harmful effects of missing data, one objective of survey researchers is to keep their prevalence as low as possible. Therefore, it is important to understand the processes that can lead to missing data.

I review the existing literature on item nonresponse and extend the satisficing model based on this review to include privacy concerns resulting in an encompassing model of item nonresponse: the probability of item nonresponse depends on the task difficulty of the item(s) divided by the product of ability and motivation of the respondent or the respondents' privacy concerns. This means that higher difficulty results in more item nonresponse and higher ability and motivation in less item nonresponse. Difficulty, ability, and motivation are separate from privacy concerns, e.g. due to item sensitivity or general mistrust. Privacy concerns are relevant when deciding whether to disclose information and do not influence the cognitive burden of retrieving the answer.

I then turn to practical strategies that could be used to decrease item nonresponse by reviewing proposed strategies. In the empirical part of the paper, I compare the effects of selected strategies. Promoting the use of showcards and translating questionnaires appear to be most promising. Both of those reduce the cognitive burden of respondents. Matching respondents' and interviewers' gender and age does not reduce item nonresponse. They could have influenced item nonresponse if respondents are more willing to share private information with interviewers similar to themselves. The results are not causal though. However, the results indicate that strategies aiming at a lower cognitive burden for respondents are our best guess to improve data quality.

This introduction follows a short theoretical discussion presenting a theoretical model for the probability of item nonresponse based on satisficing. Structured by this model, I review the literature on strategies how to reduce item nonresponse and test a selection using the \textit{European Social Survey}. The rest of the paper is devoted to this test of strategies, describing data and methods and presenting results. Finally, I summarise and discuss my review and results.

@@

Whether respondents answer a survey question and which answer they give is always a cognitive process taking place at the very moment of the interview. The \textit{survey response process} \citep{tourangeauPsychologySurveyResponse2000} involves multiple steps on behalf of the respondent. They need to comprehend the question, retrieve information from memory, eventually judge this information, map them onto the response options, and perhaps edit the response due to sensitivity or social desirability. Respondents will most likely take these steps in order but they can jump back and forth, e.g. if they need to form an opinion on the spot. But in all of these steps, item nonresponse can be introduced \citep{deleeuwPreventionTreatmentItem2003}.

The two types of item nonresponse, "Don't know" and refusal, might be related to different steps of the survey response process though. Refusals are likely introduced in the editing step when respondents do not want to answer a question although they could. They may find certain information to be too sensitive or they may not feel comfortable sharing it with the interviewer due to a lack of trust. DK is likely as an answer when the respondent cannot give a substantive answer. Either the respondent cannot answer because they do not know about the content of the question or are unable to remember an event \citep{beattyAnswerNotAnwer2002, turnerCanResponseLatencies2015}. In this situation, DK is a valid response and does not constitute a problem for data quality.
On the other hand, they might not see value in putting in the effort to give an optimal response and \textit{satisfice} \citep{krosnickResponseStrategiesCoping1991}. Satisficing refers to various shortcuts (heuristics) that survey respondents can take when answering questions. One of these shortcuts is item nonresponse. The data collected in this case are of lower quality.

However, \citet{shoemakerItemNonresponseDistinguishing2002}  have
shown that higher mental effort is related to more refusals as well. And
conversely, it is plausible that DK is used as a more polite way to refuse.
In the following sections, I will therefore not distinguish between refusal and DK even though they may have varying strengths of predictors \citep{silberEffectsQuestionRespondent2021}. Similarly, I will not consider the unproblematic case of DK as a genuine answer although differentiating between the two meanings of DK might be relevant for substantive analyses.

In the continuation of this section, I will discuss theoretical concepts that impact the likelihood of item nonresponse in the cognitive process of response formation. Later, I will combine these concepts into a theoretical model based on satisficing \citep{krosnickResponseStrategiesCoping1991}.

When it comes to item nonresponse, the key concept is the \textit{ability} of the respondents to carry out cognitive tasks. Differences in cognitive abilities are the main explanation for differences in item nonresponse across education, age, and health \citep{colsherDataQualityAge1989, pickeryImpactRespondentInterviewer1998, deleeuwPreventionTreatmentItem2003, messerDeterminantsItemNonresponse2012, silberEffectsQuestionRespondent2021}. Ethnic minorities tend to have a higher rate of item nonresponse likely caused by lower literacy and worse command of the majority language \citep{kupekDeterminantsItemNonresponse1998, pickeryImpactRespondentInterviewer1998}. \citet{meitingerPowerCultureItem2020} conclude that item nonresponse reflects broader social inequalities in abilities and access to information. The ability hypothesis is directly supported by correlations between item nonresponse and measures of intelligence \citep{hedengrenDogThatDidn2012}.

The second relevant concept is \textit{task difficulty}. When questions are more difficult or unclear, they tend to have higher rates of nonresponse \citep{holbrookImpactQuestionRespondent2006, messerDeterminantsItemNonresponse2012, holbrookInterviewerErrorsHelp2016, olsonEffectsRespondentQuestion2019}. Demographic questions are usually easier for respondents to remember, resulting in lower rates of item nonresponse compared to attitudinal and behavioral questions, which may require respondents to formulate an answer on the spot \citep{olsonEffectsRespondentQuestion2019, silberEffectsQuestionRespondent2021}.

Even if people can complete a task, they may not want to do it unless they feel that the effort is worthwhile. They need to have the \textit{motivation} to provide an optimal response. That explains why people who are more interested in the topic of a survey are less likely to leave items unanswered \citep{kochItemNonresponseEuropean2009, silberEffectsQuestionRespondent2021}. Item nonresponse is linked to conscientiousness measures as well \citep{hedengrenDogThatDidn2012}.

The editing process can also be influenced by motivation. For instance, respondents and interviewers may choose not to answer screening and filtering questions on purpose to lessen the survey workload \citep{tourangeauMotivatedMisreportingShaping2015}. This statement pertains only to data collections where the respondents know or can guess which questions serve as filters though.

When editing an answer, respondents may have concerns about their \textit{privacy}\footnote{ I use this label to encompass overall privacy concerns related to the survey, such as data processing and anonymity, as well as the desire to avoid answering specific sensitive questions}. Will the interviewer judge me if I answer truthfully? Can I trust that my data will be kept secure and confidential? This is a particular problem for questions perceived as intrusive \citep{tourangeauSensitiveQuestionsSurveys2007} like questions on income \citep{yanTrendsIncomeNonresponse2010} and sexual behaviour \citep{kupekDeterminantsItemNonresponse1998}, which often show particularly high levels of item nonresponse. However, when it comes to attitude questions about controversial political issues such as immigration, there tends to be more item nonresponse as well \citep{piekutSurveyNonresponseAttitudes2021}. Item nonresponse is indeed frequently used as a measure of question sensitivity \citep{tourangeauSensitiveQuestionsSurveys2007}. Respondents will likely have such privacy concerns immediately when they hear a sensitive question and jump from the comprehension stage to the editing stage in the survey response process \citep{tourangeauPsychologySurveyResponse2000}. They probably refuse to answer before an honest answer has been formed.  Increased item nonresponse is associated with reluctance and skepticism towards surveys and science, general privacy concerns, and mistrust \citep{silberEffectsQuestionRespondent2021}.

@@

Based on the reviewed literature, I have identified four concepts that affect the probability of item nonresponse in surveys: cognitive ability, task difficulty, motivation, and privacy concerns. However, as \citet{krosnickResponseStrategiesCoping1991} already hypothesised, these concepts are interrelated in their effect on item nonresponse. Very easy questions can be answered by less able respondents and very hard questions might even cause the most able to struggle. The resulting fraction of difficulty by ability represents the relative mental effort to answer a question. And a highly motivated respondent answers even difficult questions. \citet[225]{krosnickResponseStrategiesCoping1991} formalised the probability of satisficing.

Item nonresponse is such a satisficing strategy. Additionally, higher privacy concerns lead to more item nonresponse as well. Since this relates to another step in the survey response process, namely editing rather than comprehension, retrieval or judging, I postulate it to be independent from the other concepts.

For a complete theoretical model of harmful item nonresponse, privacy concerns therefore need to be added to the model by \citet{krosnickResponseStrategiesCoping1991}. As these concepts are (partially) interrelated and have a nonlinear relationship to item nonresponse, it is useful to formalize and summarize their relationship as follows:

$$ P_{INR} \ = f\left(\ max \left( \frac{difficulty}{ability \ * \ motivation} \ , \ privacy \right)\right) $$

\textit{The probability of an ingenuine nonsubstantive answer on behalf of the respondent is a function of the task difficulty divided by the ability and motivation of the respondent or the respondents' privacy concerns, whichever is higher.}

Please note that this theoretical model is not able to and not intended to predict the probability of item nonresponse in a given item, as highlighted by the fact that it is an undefined function. Therefore, the individual concepts do not require measurement. The maximum function emphasizes that there are two independent mechanisms, and only the dominant one will impact item nonresponse at a time. This theoretical model specifically addresses item nonresponse for a single item but its meaning is adaptable to every level of a survey.

@@

With these four concepts in mind, we can develop strategies to reduce item nonresponse and ensure better data quality in our surveys. Some of the following strategies may seem obvious and are already established standards in survey design not only because of their potential relationship to item nonresponse but to ensure the quality of substantial answers as well. Others might reduce item nonresponse but they could have negative consequences for other parts of total survey error, the combined effect of all error sources in a survey \citep{grovesTotalSurveyError2010}. They require a trade-off before implementation.

I have structured this review of strategies to reduce item nonresponse by the respective concepts they target.

@@

The level of difficulty of a task is largely determined by how the questions are designed and what type of answer is expected. To make tasks easier, it is recommended to ask short, straightforward questions that avoid any confusion or unclear concepts. Asking respondents to complete multiple tasks at once should also be avoided. For a more thorough guide on how to design questions and questionnaires, see e.g. \citet{smythDesigningQuestionsQuestionnaires2016}. The difficulty of a task is related to the type of question as well. Questions that are open-ended or allow for multiple options and ordering of categories are more likely to result in higher nonresponse rates than closed single choice items \citep{schumanOpenClosedQuestion1979, holbrookImpactQuestionRespondent2006, holbrookInterviewerErrorsHelp2016, olsonEffectsRespondentQuestion2019, silberEffectsQuestionRespondent2021}. To make it easier for respondents, visual aids like images or showcards can be used. Showcards eliminate the need for respondents to recall all response categories while answering a question. However, there is limited research on how showcards affect item nonresponse. According to a study by \citet{holbrookInterviewerErrorsHelp2016}, using showcards in survey questions led to more unanswered items. However, this may have been because showcards were only used for more challenging questions. In the European Social Survey (ESS), showcards do not appear to impact the distribution of meaningful responses in survey experiments, as noted by \citep{jackleAssessingEffectData2010}, although they did not investigate item nonresponse.

How question design affects levels of item nonresponse is very well understood and differences between questions constitute the largest part of the variance in item nonresponse \citep{olsonEffectsRespondentQuestion2019}. This highlights the importance of the single question for overall data quality.

It is important to design the entire questionnaire as simply as possible, not just the individual questions. Questionnaires that include changes in response scales, routing, and filtering tend to result in higher rates of nonresponse \citep{messerDeterminantsItemNonresponse2012}. However, routing and filtering should not increase difficulty in computer-assisted modes. Grouping questions by topic could reduce the required mental effort and item nonresponse but it also increases the likelihood of non-differentiation between items \citep{krosnickResponseStrategiesCoping1991}. Explicitly offering DK and refusal options can increase their use, as it makes respondents more aware of the possibility of an "easy way out" \citep{schumanAssessmentNoOpinion1979, beattyAnswerNotAnwer2002}.

To reduce task difficulty for members of language minorities, the questionnaire can be translated so that respondents can take the interview in the language they are most proficient in. But translating questionnaires can be costly and may affect the comparability of cases. For a review on comparability in cross-cultural surveys, see e.g. \citet{behrTranslationMeasurementInstruments2016}.

To enhance the quality of survey design, identify any errors, and ensure that respondents can complete the required tasks, it is recommended to thoroughly review the questionnaire and its implementation for data collection. Common methods for doing so include conducting reviews and pilot studies.

@@

While we cannot alter the general cognitive ability of our respondents, we can influence their ability to answer survey questions at the time of participation. To ensure a productive interview, it is important to choose an environment that encourages focus and clear communication. If possible, opt for quiet and not distracting locations at appropriate times. Having other people present during an interview can be distracting, but there is no conclusive evidence to support this claim \citep{kupekDeterminantsItemNonresponse1998, tuSocialDistanceRespondent2007, silberEffectsQuestionRespondent2021}. Respondents may become fatigued during lengthy interviews \citep{holbrookInterviewerErrorsHelp2016, olsonEffectsRespondentQuestion2019}.

@@

Motivation could decrease throughout the interview as well. While web surveys have used different page layouts and progress bars to combat this issue, the effectiveness of these methods is uncertain \citep{peytchevWebSurveyDesign2006, yanShouldStayShould2011, sarrafSurveyPageLength2014}. The research on cooperation enhancement, such as through incentives, has mainly focused on unit nonresponse. However, some of these methods could also be effective in increasing item nonresponse. After all, unit and item nonresponse are linked: respondents that initially refused to participate have higher levels of item nonresponse \citep{yanRelationUnitNonresponse2010, frickerExaminingRelationshipNonresponse2010}.

@@

Survey researchers should address privacy concerns to encourage respondents to answer by ensuring the security and anonymity of their data. It is important to communicate why the data is collected, how it will be processed, and how privacy is protected. This is not only ethically advisable but also often a legal requirement.

When conducting face-to-face surveys, the trust between the respondent and interviewer is influenced by their relationship. Scholars have hypothesized that respondents are more likely to trust interviewers who they perceive to be similar to themselves. To test this hypothesis, studies have been conducted to examine the impact of matching characteristics between the respondent and interviewer. \citet{vercruyssenEffectSociodemographicMis2017} observe less item nonresponse when interviewers and respondents are matched in age. Additionally, matching gender reduces item nonresponse for males but increases it for females. \citet{piekutSurveyNonresponseAttitudes2021} found female interviewers experienced higher rates of nonresponse but there was no significant correlation between the gender of the interviewer and the gender of the respondent. \citet{silberEffectsQuestionRespondent2021} found that education matching has no effect while \citet{tuSocialDistanceRespondent2007} find age and education matching to be potent predictors of item nonresponse. A test that could be interesting to conduct is whether pairing interviewers and respondents who share the same immigration status and/or ethnicity would make a difference. Immigrants tend to have higher levels of item nonresponse, language barriers, and I could imagine that some of them may be mistrustful towards interviewers due to racist experiences and a fear of discrimination.

@@

So far, this article has reviewed and summarized the literature on item nonresponse in surveys. I have suggested a theoretical model, an extension of the satisficing model by \citet{krosnickResponseStrategiesCoping1991}, as a conceptual summary that can inform our survey design and I have reviewed strategies to reduce item nonresponse and categorized them accordingly. In the remainder of the article, I am going to test a few selected strategies to reduce item nonresponse derived from the theoretical model and the literature review. All of these strategies could change at least one of the four concepts from the theoretical model and therefore could have a causal connection to item nonresponse. Whether these strategies actually do change the associated concepts and how strongly their effect translates into changes in item nonresponse will be central to my empirical analysis.

\begin{enumerate}
 \item During an interview, respondents may experience a decrease in concentration and motivation to answer questions as time goes on. As a result, item nonresponse may become more common the longer the interview lasts. To maintain high data quality and reduce item nonresponse, it might be advisable to keep questionnaires as short as possible.

 \item It is likely that interferences and the presence of others during an interview can cause item nonresponse, as they may distract the respondent and make them hesitant to answer certain questions in front of people they know. As a result, it might be beneficial that interviews are conducted without other people present, if feasible.

 \item To make answering easier for respondents, showcards could be provided so they do not have to remember the response scale. Showcards would then lower the required cognitive effort and reduce item nonresponse.

 \item Respondents who primarily speak a different language at home may face difficulties in the response process. To ensure data quality from these respondents, one option is to translate the questionnaire, although this can be costly and may present comparability problems.

 \item According to previous studies, people may feel more comfortable answering questions from interviewers who share similar social characteristics, such as gender and age. If true, survey agencies could assign interviewers based on demographic information if it is available in the sampling frame.
\end{enumerate}

Table 1 summarises the selected strategies that I am going to test in my empirical analysis. The second column shows which concepts play a role in the hypothesized mechanism linking the respective strategy to item nonresponse. The third column gives the expected direction of the relationship between strategy and item nonresponse, e.g. the longer the questionnaire, the more item nonresponse. These are also the expected signs of the coefficients if the strategies work as imagined.

\begin{table}[h]
\caption{Reduction Strategies to be Tested}
\centering
\begin{tabular}{| l l l |}
\hline
Strategy & Mechanism & Expectation \\
\hline
Length of the Questionnaire & Ability, Motivation & positive \\
Interference of the Interview & Ability, Privacy & positive \\
Use of Showcards & Difficulty & negative \\
Interview not primary Language & Difficulty & positive \\
Gender Matching & Privacy & negative \\
Interviewer more than 10 years older & Privacy & positive \\
Respondent more than 10 years older & Privacy & positive \\
\hline
\end{tabular}
\end{table}

@@

To test the effectiveness of some potential strategies to reduce item nonresponse, I use the \textit{European Social Survey (ESS)} Round 9 collected between August 2018 and January 2020 \citep{essericEuropeanSocialSurvey2019}. The ESS is a biannual face-to-face trend survey on attitudes and beliefs towards social and political topics in Europe established in 2001. In each country and round, the ESS draws a new random sample of the residential population of 15 years and older aiming for a minimum response rate of 70\%. Most countries use computer-assisted personal interviews for data collection and the questionnaire is designed to take about one hour. The data release 3.1 includes data from 49,519 respondents from 29 countries. For more information on the data, see the supplementary material or visit \href{https://www.europeansocialsurvey.org}{europeansocialsurvey.org}.

@@

The three dependent variables are the sum of DK, the sum of refusals, and the total sum of item nonresponse for every respondent. Non-responses are only counted for variables that are presented to all respondents and not affected by filtering questions. Respondents are not given the option to respond with DK or refusal, but interviewers are instructed to record them explicitly and without further probing. It is up to the interviewer to interpret a non-response as either a refusal or DK.
